---
title: "HW1_RJS_17208987"
author: "Liang Tong"
date: today
format: beamer
---

## 1.Use `data.table` to read in the data and assign the correct class to the variables.

The data I choose is:

1.  ***Human Development Indicators for Ireland***

2.  ***Human Development Indicators for China***

In this step I'll load required libraries and datasets

```{r 1}
# load library
library(data.table)
library(ggplot2)

# load datasets, skip first row
country1 <- fread("Human Development Indicators for Ireland2024.csv",skip=1)
country2 <- fread("Human Development Indicators for China2024.csv",skip=1)


# convert to data table format, good for large size
country1_DT <- as.data.table(country1)
country2_DT <- as.data.table(country2)
class(country1_DT$'#indicator+value+num')
```

Assign the correct class to the variables, here the variable unit is correct: *numeric* for **'Value'** and *int* for **'Year'**

```{r 2}
#reset names and check unit
 setnames(country1_DT, c("country_code", "country_name", "indicator_id", "indicator_name", "index_id", "index_name", "value", "year"))
  setnames(country2_DT, c("country_code", "country_name", "indicator_id", "indicator_name", "index_id", "index_name", "value", "year"))
  country1_DT
```

## 2. Merge the data data sets using `data.table`

```{r 3}
# base R example， by=(); for data.table , .on(...)
#print(colnames(country2_DT))

# useing data table to joining tables
merged_DT <- rbind(country1_DT, country2_DT) 
#merged_DT <- country1_DT[country1_DTT, on = .(country_code,country_name, #indicator_id,indicator_name, index_id,index_name, value, year)]
merged_DT
```

## 3. Quick data exploration

here show the variable names of the merged data sets:

-   we quickly explore the difference between Ireland and China

-   The index I'm looking intro is

    |     | Indicator_name                         | Indicator_id |
    |-----|----------------------------------------|--------------|
    | 1   | Expected Years of Schooling (years)    | eys          |
    | 2   | Life Expectancy at Birth (years)       | le           |
    | 3   | Material footprint per capita (tonnes) | mf           |

```{r 4}
#show the variable name
print(colnames(merged_DT))
```

Here we can see The mean,min and max for Ireland and China's **Expected Years of Schooling (years)** for past 33 years. Ireland has a much higher eys for the quick check.

```{r 5}
#Ireland and China's gender mean In equality Index
merged_DT[indicator_id=="eys", 
          .(eys_mean = mean(value),eys_min = min(value),eys_max = max(value), .N),
          by= country_name]
```

do the same for **Life Expectancy at Birth (years)** and **Material footprint per capita (tonnes)**. This time Ireland has a slightly higher le and a much higher mf in the past 33 years

```{r 6}
merged_DT[indicator_id=="le", 
          .(le_mean = mean(value),le_min = min(value),le_max = max(value), .N),
          by= country_name]
```

```{r 7}
merged_DT[indicator_id=="mf", 
          .(Gmf_mean = mean(value),mf_min = min(value),mf_max = max(value), .N),
          by= country_name]
```

## 4. Analysis using data.table - keyby() used

compare **Expected Years of Schooling (years)** and rank it by year and country. we could find that both countries's year is increasing over the past 33 years

```{r 8}
eys_IRL <- merged_DT[ country_name=="Ireland" & indicator_id=="eys",
                      .SD,
                      keyby=.(year,value)]
eys_CHN <- merged_DT[ country_name=="China" & indicator_id=="eys",
                      .SD,
                      keyby=.(year,value)]
eys_IRL
#eys_CHN
```

```{r 9}
eys_CHN
```

now analysis the **Life Expectancy at Birth (years).** This time get the same result but using different method-'keyby = TRUE'. sort using country list (China 33 then Ireland 33). we found that the life time is increasing , but Ireland is slightly higher than China

```{r 10}
le_both <- merged_DT[ indicator_id=="le",
                      .SD,
                      by=.(country_name,year, value),
                      keyby = TRUE]#same result
le_both
```

For **Material footprint per capita (tonnes)**, I want to check data for **recent 24 years.** We can found that Ireland has a overall higher footprint. China 's is increase and Ireland's is increase then decreasing in recent 24 years

```{r 11}
mf_both <- merged_DT[ indicator_id=="mf" & year>2000,
                      .SD,
                      by=.(country_name,year, value),
                      keyby = TRUE]
mf_both
```

## 5. Plotting

plot below comparing the **Expected Years of Schooling (in years)** between Ireland and China. The data shows that Ireland's overall schooling time is higher than China's. Notably, there is a significant jump around 1998 in Ireland. Additionally, both countries have shown an increasing trend in schooling time over the years.

```{r 12}
#x <- eys_IRL$year
#y1 <- eys_IRL$value
#y2 <- eys_CHN$value
#plot(x,y1,x,y2)
library(hrbrthemes)
ggplot(merged_DT[indicator_id=="eys"], aes(x=year, y=value, shape=country_name)) + 
    geom_point(size=6, color="lightblue") +
    ggtitle("Expected Years of Schooling (years)") +
    theme_ipsum()


```

For **Life Expectancy at Birth (years)** I want to plot the box plot. We found that Ireland has a higher mean and median

```{r 13}
ggplot(merged_DT[indicator_id=="le"], aes(x=country_name, y=value)) + 
    geom_boxplot(fill="slateblue", alpha=0.2) + 
    xlab("Country name")+
    ggtitle("Life Expectancy at Birth (years)")
```

For **Material footprint per capita (tonnes)**, I want to plot the line chat

```{r 14}
ggplot(merged_DT[indicator_id=="le"& year>2000], aes(x=year, y=value, group=country_name, color=country_name)) +
    geom_line()+
  ggtitle("Material footprint per capita (tonnes)-recent 24 years")
```
